---
  title: "Final Project"
  author: "Chenxi Du, Chuyan Luo, Danyao Yu, Hang Xu, Siqi Zhang, Yiyao Zhou, Yuwei Lin"
output: html_document
---

For this project, we would like to explore more about the volatility forecasting area and try to apply it to real world.

# Looking at the data
```{r}
library(forecast)
library(zoo)
library(xts)
library(urca)
library(ggplot2)
library(rugarch)
library(plyr)

#IEO is an oil exploitation ETF, frequently trading
IEO<-read.csv("IEO.csv")
#Make IEO time series
IEO.ts <- xts(IEO$Adj.Close, as.Date(IEO$Date))
plot(IEO.ts)
#create return and variance
ret<-diff(log(IEO.ts))
plot(ret^2)
#examine pacf of IEO time series, significant correlation with one lag. 
#Highly likely to be random walk and need further test for unit root.
Pacf(IEO.ts)
#examine volatility of return, the result is persistent and gradually decay
vol<-sqrt(ret^2)
Pacf(vol)

#create training and validation set
train.ts <- window(IEO.ts, end = "2015-12-31")
valid.ts <- window(IEO.ts, start = "2016-01-01")
ntrain <- length(train.ts)
nvalid <- length(valid.ts)
```
First we looked at the pattern of price and return. Price time series has a slightly upward trend and the Pacf indicate that it is likely that the data will have unit root.

# Some models to forecast 
```{r}
# Arima Model

#DF Test on Whole Data Set, Fail to reject null hypothesis → unit root 
df.test <- ur.df(IEO.ts,type="trend",selectlags="BIC")
print(summary(df.test))
sum <- summary(df.test)
teststat <- sum@teststat[3]
# critical value at 5 percent
critical <- sum@cval[3,2]
abs(critical)<abs(teststat) #Not reject null, series has unit root

# Building the model
ArimaMod <- auto.arima(train.ts, d=1, ic="bic", seasonal = FALSE)
summary(ArimaMod) #give us (0,1,0), that is a random walk
fcast_Arima<- Arima(valid.ts, model = ArimaMod) #one step ahead forecasting
# make fitted value both in training and validation time series data;preparing for ploting
fcast_fitted_Arima.ts <- xts(fcast_Arima$fitted, as.Date(IEO$Date[(ntrain+1):(ntrain+nvalid)]))
mod_fitted_Arima.ts <- xts(ArimaMod$fitted, as.Date(IEO$Date[1:ntrain]))
# examine model's accuracy
print(accuracy(fcast_Arima$fitted, valid.ts))

# Plot, the fitted values highly coincide with actual value due to one step ahead process
plot(IEO.ts, ylab="Price",xlab="Time",bty="l",main="Forecast of IEO Price(Arima(0,1,0))", flty=2)
lines(mod_fitted_Arima.ts, lwd=0.5,col="aquamarine3")
lines(fcast_fitted_Arima.ts, lwd=0.5,col="azure3")
```
We further examined the data and found that the time series has unit root. Based on the results, we set d=1 in the ARIMA model. The auto.arima function gives us ARIMA(0,1,0), which is a random walk.
```{r}
# Exponential Filter Forecast

# First, estimate additive trend filter
filter.mod <- ets(train.ts, model = "AAN")
# Now, build forecasts for validation periods (uses no data there)
filter.pred <- forecast(filter.mod, h = nvalid)
plot(filter.pred)
# Now, build one step ahead forecasts for validation data, using estimated model
filter.onestep <- ets(valid.ts, model = filter.mod)
# make it time series through xts
filter.onestep.fitted.ts <- xts(filter.onestep$fitted, as.Date(IEO$Date[(ntrain + 1):(ntrain + nvalid)]))
plot(IEO.ts, main = "Filter Forecast (with trend)")
lines(filter.onestep.fitted.ts,lwd = 1, col="azure3", lty = 2)
# check accuracy
print(accuracy(filter.onestep$fitted, valid.ts))
```
For the exponential filter, we chose the model with additive error, additive trend and no seasonality. We tried both multiperiod forecasting and one-step ahead forecasting for this model. It is obvious that one-step ahead forecasting does much better.
```{r}
# Naive Model

# Use Mean of full data set as Naive Forecast
nALL <- length(IEO.ts)
naive.valid <- rep(mean(IEO.ts),nvalid)
naive.ts<-xts(naive.valid, as.Date(IEO$Date[(ntrain + 1):(ntrain + nvalid)]))
#Plot
plot(IEO.ts)
lines(naive.ts,col="blue") 
```


# Model Comparison
```{r}
# RMSE of Naive, Arima and Filter
naive.valid.res <- valid.ts-naive.valid
arima.valid.res<-residuals(fcast_Arima)
filter.valid.res<-residuals(filter.onestep)
V.rmse <- rep(NA,3)
V.mae<-rep(NA,3)
V.rmse[1] <- sqrt(mean((naive.valid.res)^2, na.rm =TRUE))
V.rmse[2] <- sqrt(mean((arima.valid.res)^2, na.rm =TRUE))
V.rmse[3]  <- sqrt(mean((filter.valid.res)^2, na.rm =TRUE))
V.mae[1]<-mean(abs(naive.valid.res))
V.mae[2]<-mean(abs(arima.valid.res))
V.mae[3]<-mean(abs(filter.valid.res))
# Compare RMSE for three models 
a<-data.frame(V.rmse=V.rmse,
              V.mae=V.mae,
              row.names = c("Naive", "Arima","Filter"))
print(a)
```
The result shows that the exponential fitler model does the best with smallest RMSE and MAE due to one step ahead process. However, the RMSE and MAE for arima model is very close to it. Naive model is least accurate.
```{r}
# perform some Diebold/Mariano tests
print("Diebold/Mariano ARIMA versus Naive")
print(dm.test(residuals(fcast_Arima),naive.valid.res))

print("Diebold/Mariano Exponential Filter versus Naive")
print(dm.test(residuals(filter.onestep),naive.valid.res))

print("Diebold/Mariano ARIMA versus Exponential Filter")
print(dm.test(residuals(fcast_Arima),residuals(filter.onestep)))
```
The Diebold-Mariano test gives similar result as above -- both ARIMA model and exponential filter model are significantly better than naive model, whereas exponential filter model beats ARIMA model but p value is not significant.


```{r}
#AGG ETF
AGG<-read.csv("AGG.csv")
Return<-diff(log(AGG$Adj.Close))
Date<-AGG[-1,1]
bond_return.ts <- zoo(Return,as.Date(Date))
bond_return.valid <- window(bond_return.ts,start=as.Date("2016-01-01"),end = as.Date("2018-12-04"))
tail(bond_return.valid)
#compare variance of IEO and bond
IEO_var<-window(ret^2,start=as.Date("2016-01-01"),end = as.Date("2018-12-04"))
bond_var<-bond_return.valid^2
plot(bond_var)
#Plot both varainces
ggplot() + 
  geom_line(aes(y = IEO_var, x = time(IEO_var), col = "IEO's variance"))+
  geom_line(aes(y = bond_var, x = time(bond_var), col = "Bond's variance")) +
  labs(title ="Bond's varaince vs IEO's varaince", x = "Time", y = "Daily Variance")+
  theme(legend.position = "bottom")
# The volatility of bond respecting to IEO is very small

```

```{r}
#Aggregation Function
monagg <- function(data)
{
  temp <- tail(data,n=1)
  m <- nrow(data)
  # these are the three lines to really pay attention to
  temp$ret <- prod(1+data$V1)-1
  temp$m <- m # number of days
  temp$sd<-sd(data$V1)
  # return monthly record
  temp
}
```
#Volatility forecast using GARCH(1,1)
As summarized from the previous part of the project, stock price and stock return are hard to forecast because most of the times they are random walk. However, the volatility of stock return presents a strong correlation through time and can be used to forecast the future performance of stock. The GARCH models are good for stock volatility foracasting because it incorporates long historical data with current data added every forecast period. The following chunk shows the volatility forecast of the IEO ETF using GARCH(1,1).
```{r}
#Generate IEO's Returns
IEO<-read.csv("IEO.csv")
IEO.ts <- xts(IEO$Adj.Close, as.Date(IEO$Date))
rsp <- diff(log(IEO.ts))
rsp<-rsp[-1,]
head(rsp)
train.ts <- window(rsp, end = "2015-12-31")
valid.ts <- window(rsp, start = "2016-01-01")
ntrain <- length(train.ts)
nvalid <- length(valid.ts)
# GARCH Model in Training set
spec <- ugarchspec(variance.model=list(model="sGARCH",garchOrder=c(1,1)),
                   mean.model=list(include.mean=T,armaOrder=c(0,0)))
fittrain <- ugarchfit(spec = spec, data=train.ts)
# Forecasting Volatility in Validation set
setfixed(spec) <- as.list(coef(fittrain))
ugarchfilter <- ugarchfilter(spec=spec,data=valid.ts)
# fitted volatility
garchVolvalid.ts <- zoo(sigma(ugarchfilter))
#check mean of fitted volatility
mean(garchVolvalid.ts)
#plot fitted volatility
ggplot() + 
  geom_line(aes(y = sqrt(IEO_var), x = time(IEO_var), col = "IEO's volatility"))+
  geom_line(aes(y = garchVolvalid.ts, x = time(garchVolvalid.ts), col = "fitted volatility")) +
  labs(title ="Garch fitted volatility vs actual volatility", x = "Time", y = "Daily Volatility")+
  theme(legend.position = "bottom")
```
As shown above, we partitioned the daily return data of IEO into training and validation set. The split point is 2015-12-31 with lengths of training/validation set around 3:1. We used the training set to train the GARCH(1,1) model and applied the model to forecast the volatility in the validation set. The graph shows the comparason of the actual daily volatility and the forecast volatility of IEO in the validation set. As we can see, the GARCH(1,1) model can nicely capture the trend of the volatility with much smoother curve compared with the actual one. The smoothing effect is because GARCH(1,1) incorporates more historical data hence less exposed to sharp changes in the current data. We believe this is a rather good feature for our strategy because it helps us avoid the sharp change in our asset weight only due to temporary fluctuation, so we can focus more on the real shift in the volatility regime of our assets.
```{r}
#Build Strategy:
#set for 1 percent daily volatility target
#use GARCH fitted volatility as estimate of next day's volatility to adjust daily weight
#use Arima fitted return as estimate of next day's return
#compare dynamic strategy to constant strategy and whole equity strategy through monthly aggregation 
target<-0.01
weight <- target/garchVolvalid.ts
#The maximum weight is 1
weight[weight$x>1]=1
length(weight)
# constant portfolio benchmark
mweight <- mean(weight)

# dynamic daily portfolio 
#one step ahead fitted return
IEO_fitted.return<-diff(log(fcast_Arima$fitted))
head(IEO_fitted.return)
length(IEO_fitted.return)
#Adjust length the same as fitted return
weight<-weight[-1]
length(weight)
bond_return.valid<-bond_return.valid[-1]
length(bond_return.valid)

#Return for dynamic portfolio
pret  <-  as.vector(IEO_fitted.return*weight)  + as.vector(bond_return.valid*(1-weight))
#Return for constant weight portfolio
pretConstant <-  as.vector(IEO_fitted.return)*mweight  + as.vector(bond_return.valid)*(1-mweight)
#Return for Whole IEO ETF
pretEquity   <- as.vector(IEO_fitted.return)

# Monthly Aggregate Return for Dynamic Strategy
pret.ts <- xts(pret, as.Date(IEO$Date[(ntrain+2):(ntrain+nvalid)]))
pret.month <- as.data.frame(pret.ts)
pret.month$date <- rownames(pret.month)
pret.month$key <-as.numeric(format(as.Date(pret.month$date),"%Y"))*100+as.numeric(format(as.Date(pret.month$date),"%m"))
pret.month<-pret.month[-1,]

pret.month <- ddply(pret.month,.variables = "key",.fun = monagg)
#mean annualized return
mean(pret.month$ret)*12
#mean annualized standard deviation
mean(pret.month$sd,na.rm = TRUE)*sqrt(12)


# Monthly Aggregate Return for Constant Strategy
pretConstant.ts <- xts(pretConstant, as.Date(IEO$Date[(ntrain+2):(ntrain+nvalid)]))
pretConstant.month <- as.data.frame(pretConstant.ts)
pretConstant.month$date <- rownames(pretConstant.month)
pretConstant.month$key <-as.numeric(format(as.Date(pretConstant.month$date),"%Y"))*100+as.numeric(format(as.Date(pretConstant.month$date),"%m"))
pretConstant.month<-pretConstant.month[-1,]
pretConstant.month <- ddply(pretConstant.month,.variables = "key",.fun = monagg)
#mean annualized return
mean(pretConstant.month$ret)*12
#mean annualized standard deviation
mean(pretConstant.month$sd,na.rm = TRUE)*sqrt(12)

# Monthly Aggregate Return for Whole ETF
pretEquity.ts <- xts(pretEquity, as.Date(IEO$Date[(ntrain+2):(ntrain+nvalid)]))
pretEquity.month <- as.data.frame(pretEquity.ts)
pretEquity.month$date <- rownames(pretEquity.month)
pretEquity.month$key <-as.numeric(format(as.Date(pretEquity.month$date),"%Y"))*100+as.numeric(format(as.Date(pretEquity.month$date),"%m"))
pretEquity.month<-pretEquity.month[-1,]
pretEquity.month <- ddply(pretEquity.month,.variables = "key",.fun = monagg)
#mean annualized return
mean(pretEquity.month$ret)*12
#mean annualized standard deviation
mean(pretEquity.month$sd,na.rm = TRUE)*sqrt(12)

# Compare annualized mean, standard deviation and sharp ratio for three strategies 
A.return <- rep(NA,3)
A.volatility<-rep(NA,3)
A.sharpe_ratio<-rep(NA,3)
A.return[1] <- mean(pret.month$ret)*12
A.return[2] <- mean(pretConstant.month$ret)*12
A.return[3]  <- mean(pretEquity.month$ret)*12
A.volatility[1]<-mean(pret.month$sd,na.rm = TRUE)*sqrt(12)
A.volatility[2]<-mean(pretConstant.month$sd,na.rm = TRUE)*sqrt(12)
A.volatility[3]<-mean(pretEquity.month$sd,na.rm = TRUE)*sqrt(12)
A.sharpe_ratio[1]<-(mean(pret.month$ret)*12-0.03)/(mean(pret.month$sd,na.rm = TRUE)*sqrt(12))
A.sharpe_ratio[2]<-(mean(pretConstant.month$ret)*12-0.03)/(mean(pretConstant.month$sd,na.rm = TRUE)*sqrt(12))
A.sharp_ratio[3]<-(mean(pretEquity.month$ret)*12-0.03)/(mean(pretEquity.month$sd,na.rm = TRUE)*sqrt(12))


a<-data.frame(A.return=A.return,
              A.volatility=A.volatility,
              A.sharpe_ratio=A.sharp_ratio,
           row.names = c("Dynamic Strategy", "Constant Strategy","Whole Equity Strategy"))
print(a)

```
Above is a graph shows out of sample one step ahead GARCH volatility forecast.

# Dynamic Strategy
We built a portfolio using two ETF: iShares US Oil & Gas Explor & Prod ETF(IEO), and iShares Core U.S. Aggregate Bond ETF(AGG).
In this strategy, GARCH fitted volatility was used as an estimate of next day's volatility to adjust the daily weight. We set 1 percent as the daily volatility target. To be more specific, the weight of the IEO ETF was calculated using the ratio of the volatility target to the out of sample one step ahead GARCH volatility estimated value. And the weights of the assets was rebalanced on a daily frequency. Additionally, we added a constraint that the maximum weight of IEO is one, which means short is not allowed in this portfolio.

To evaluate the performance of the dynamic strategy, we created a monthly aggregation function to calculate the monthly return and monthly volatility.
The dynamic strategy achieved a 5.175% annualized return in the validation period (from 2016-01-01), and the Sharpe ratio is 0.6884. The Dynamic strategy shows a better performance compared with the constant strategy and the whole equity strategy: its Sharpe ratio is 99.42% higher than the constant strategy and 53.46% higher than the whole equity strategy.

# Conclusion
1. Price of IEO ETF is a random walk. ARIMA(0,1,0) model and Exponential Filter with trend can both do a significantly better job in forecasting the price one step forward compared with the Naive model, while ARIMA(0,1,0) slightly ourperforms Exponential Filter in terms of RMSE.

2. While stock price is random and hard to forecast, stock volatility has strong autocorrelation, so the forecasting of volatility seems to make more sense. The GARCH(1,1) model does a nice job in forecasting the daily volatility of IEO ETF in terms of capturing its trend with smoothing effect. The forecasted volatility is the base of our strategy construction.

3. To balance the volatility of IEO and hence gain the best Sharpe ratio, we add a bond ETF, AGG, which is all made up of US Treasury bonds and investment grade corporate bonds, to our portfolio. When IEO volality is forecasted higher than "target", we reduce the weight of IEO and raise the weight of AGG, vice versa, with daily balance. The result turns out satifactory with annualized Sharpe ratio of 0.6884, 99.42% higher than the constant strategy and 53.46% higher than the whole equity strategy.

# Improvements 
1. With different cutting point of training and validation set, there can be different models, so recursive or rolling method could be better.

2. We only used GARCH(1,1) in our project but we could actually try other models make some robust analysis to make sure we find a best model.

3. We didn't find a better way to get target volatility. Maybe we could make deeper analysis into the volatility regimes of our assets and find a more flexible and scientific way to set the target.

4. When calculating the future weights of IEO, we divided "target" only by the forecasted volatility of IEO, because we believe the correlation of IEO and AGG is near 0. However, to get a more unbiased weight vector, we need to forecast the covariance matrix of the whole portfolio as well.
